Polynomial phase estimation by least squares phase unwrapping

### 2013-06-29 paper.tex
minor
added to pps, dumb tables

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  The LSU estimator can be computed in a reasonable amount of time for data sets of moderate size using existing general purpose algorithms from algebraic number theory.  Under mild conditions on the distribution of the noise we describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed.  No `high SNR' assumption (common in the literature) is required in our analysis.  As a result the LSU estimator is accurate, even for very noisy signals, if the number of observations is sufficiently large.  A key feature is that the LSU estimator is accurate over a far wider range of parameters than many popular existing estimators.  Monte-Carlo simulations support our theoretical results and demonstrate the excellent statistical performance of the LSU estimator when compared with existing state-of-the-art estimators.
\end{abstract}

@@

%\bibliographystyle{unsrt}

@@

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  
A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In~\cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999,Agrell2002} that exactly computes a nearest lattice point, and \emph{Babai's nearest plane algorithm}~\cite{Babai1986} that only produces an approximation. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for $N$ less than about $60$.  Babai's algorithm is computationally simple, but its statistical performance has been observed to be comparatively poor at low signal to noise ratio (SNR)~\cite{McKilliam2009asilomar_polyest_lattice}. Of interest is that the lattices considered have significant structure, and may admit fast nearest point algorithms.  This has been studied in~\cite{McKilliam2010thesis} where polynomial time algorithms were found that compute the nearest point exactly.  Although polynomial time, the algorithms are still computationally demanding in practice.  Nevertheless fast (exact or approximate) algorithms may exist for these special lattices, that are yet to be discovered.  In this paper we apply another general purpose approximate nearest point algorithm called the $K$-best algorithm~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} for approximating the LSU estimator.  We show that this provides near sphere decoder (exact) performance, but can be computed in a reasonable amount of time if $N$ is not too large (approximately $N$ less than $1000$).  %The $K$-best method is motivated by sequential decoding algorithms and, in the context of polynomial phase estimation, bares some resemblance to the Bayesian unwrapping algorithm studied by Morelande~\cite{Morelande_bayes_unwrapping_2009_tsp}.
%We show by Montecarlo simulation that the LSU estimator approximated using the $K$-best method is statistically superior to state-of-the-art estimators from the literature including the HAF and CPF.  We also compare the computational performance of the estimators and find that BLERG.

@@

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise process $\{X_n\}$. The proof of strong consistency is given in Section~\ref{sec:strongconstproof}.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of this is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the LSU estimator with existing estimators including the HAF, PHAF and CPF.  These simulations agree with the asymptotic properties.

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find a lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, find $\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.$  The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself is considered as an additional input parameter \cite{micciancio_hardness_2001}.  On the other hand, fast algorithms exist for specific lattices~\cite{SPLAG,Conway1982FastQuantDec,McKilliam2009CoxeterLattices,McKilliam2008}.  Even in the general case, where the lattice has no known structure, there are algorithms that compute a nearest lattice point in reasonable time if the dimension is small~\cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. These algorithms typically operate using a tree search to enumerate all lattice points inside a ball of radius sufficiently large to contain a nearest point.  The algorithms vary in the order that branches in the tree are searched.  A popular approach due to Schnorr and Euchner~\cite{schnorr_euchner_sd_1994,Agrell2002} searches branches in a greedy order, and this is the approach we implement here.  We refer to this algorithm as the~\emph{sphere decoder} in this paper, although other variants of the algorithm also go by this name.  With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.

Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Another pragmatic approximate approach is to employ the sphere decoder, but prune branches in the search tree to ensure that there are at most a finite number, say $K$, remaining branches at any time.  This approach has been studied previously~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} and resembles what is called the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} from coding theory.  The approaches vary in how branches are pruned.  In our implementation we prune branches using the same greedy metric as that used by the Schnorr and Euchner sphere decoder~\cite{schnorr_euchner_sd_1994,Agrell2002} for deciding the order in which branches are searched.  Following~\cite{Zhan2006_K_best_sphere_decoder} we refer to this as the \emph{$K$-best algorithm}.  The $K$-best algorithm requires $O(NK^2 \log K)$ arithmetic operations. The maximum number of branches $K$ is free to be chosen.  For polynomial phase estimation we have found that setting $K=10N$ yields excellent results, and in this case the algorithm requires $O(N^3 \log N)$ operations.  Simulations in Section~\ref{sec:simulations} show that this algorithm produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for $N$ less than about 1000.

@@

Figure~\ref{plot:bench} shows the running time of the LSU estimator computed using the $K$-best method versus the HAF estimator for polynomial phase signals of order 3.  The HAF estimator is computed using the fast Fourier transform and has complexity $O(N \log N)$.   As described in Section~\ref{sec:lattice-theory} the LSU estimator computed using the $K$-best method requires in the worst case $O(N^3 \log N)$ operations.  The $K$-best method tends to be faster when the SNR is large.  This is due to the algorithm rapidly finding a nearest point, and subsequently terminating branches in its search tree early.  On the other hand, when the SNR is small, all branches in the tree tend to be searched.  Interestingly the LSU estimator computed using the $K$-best algorithm is comparable with the HAF when the SNR is large and when $N$ is approximately less than 100.  The LSU estimator can be computed in a reasonable amount of time (a few minutes) when $N=1000$.  The algorithm becomes prohibitively expensive for large $N$.  Benchmarks have been performed with polynomial phase signals of order larger than $3$.  The results are omitted due to space constraints.  The running time of the LSU estimator does not appear to change significantly as $m$ increases.  This is expected since a nearest point in a lattice of approximately the same dimension $N-m-1$ must be computed.  The complexity of the HAF estimator grows linearly with $m$.

@@

\nocite{OShea_refiningpps_2010}

>>>>

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  The LSU estimator can be computed in a reasonable amount of time for data sets of moderate size using existing general purpose algorithms from algebraic number theory.  Under mild conditions on the distribution of the noise we describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed.  %No `high SNR' assumption (common in the literature) is required in our analysis.  As a result the LSU estimator is accurate, even for very noisy signals, if the number of observations is sufficiently large.  
A key feature is that the LSU estimator is accurate over a far wider range of parameters than many popular existing estimators.  Monte-Carlo simulations support our theoretical results and demonstrate the excellent statistical performance of the LSU estimator when compared with existing state-of-the-art estimators.
\end{abstract}

@@

% \bibliographystyle{unsrt}

@@

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam2010thesis,McKilliam_polyphase_est_icassp_2011}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  
A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In~\cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999,Agrell2002} that exactly computes a nearest lattice point, and \emph{Babai's nearest plane algorithm}~\cite{Babai1986} that only produces an approximation. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for $N$ less than about $60$.  Babai's algorithm is computationally simple, but its statistical performance has been observed to be comparatively poor at low signal to noise ratio (SNR)~\cite{McKilliam2009asilomar_polyest_lattice}. Of interest is that the lattices considered have significant structure, and may admit fast nearest point algorithms.  This has been studied in~\cite{McKilliam2010thesis} where polynomial time algorithms were found that compute the nearest point exactly.  Although polynomial time, the algorithms are still computationally demanding in practice.  Fast (exact or approximate) algorithms may exist for these special lattices, that are yet to be discovered.  In this paper we apply another general purpose approximate nearest point algorithm called the $K$-best algorithm~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} for approximating the LSU estimator.  We show that this provides near sphere decoder (exact) performance, but can be computed in a reasonable amount of time if $N$ is not too large (approximately $N$ less than $1000$).  %The $K$-best method is motivated by sequential decoding algorithms and, in the context of polynomial phase estimation, bares some resemblance to the Bayesian unwrapping algorithm studied by Morelande~\cite{Morelande_bayes_unwrapping_2009_tsp}.
%We show by Montecarlo simulation that the LSU estimator approximated using the $K$-best method is statistically superior to state-of-the-art estimators from the literature including the HAF and CPF.  We also compare the computational performance of the estimators and find that BLERG.

@@

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise process $\{X_n\}$. The proof of strong consistency is given in the appendix.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of this is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the LSU estimator with existing estimators including the HAF, PHAF and CPF.  These simulations agree with the asymptotic properties.

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find a lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, find $\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.$  The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself is considered as an additional input parameter \cite{micciancio_hardness_2001}.  On the other hand, fast algorithms exist for specific lattices~\cite{SPLAG,Conway1982FastQuantDec,McKilliam2009CoxeterLattices,McKilliam2008}.  Even in the general case, where the lattice has no known structure, there are algorithms that compute a nearest lattice point in reasonable time if the dimension is small~\cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. These algorithms typically operate using a tree search to enumerate all lattice points inside a ball of radius sufficiently large to contain a nearest point.  The algorithms vary in the order that branches in the tree are searched.  A popular approach due to Schnorr and Euchner~\cite{schnorr_euchner_sd_1994,Agrell2002} searches branches in a greedy order.  We refer to this algorithm as the~\emph{sphere decoder} in this paper, although other variants of the algorithm also go by this name.  With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.

Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Another pragmatic approximate approach is to employ the sphere decoder, but prune branches in the search tree to ensure that there are at most a finite number, say $K$, remaining branches at any time.  This approach has been studied previously~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} and resembles what is called the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} from coding theory.  The approaches vary in how branches are pruned.  In our implementation we prune branches using the same greedy metric as that used by the Schnorr and Euchner sphere decoder~\cite{schnorr_euchner_sd_1994,Agrell2002} for deciding the order in which branches are searched.  Following~\cite{Zhan2006_K_best_sphere_decoder} we refer to this as the \emph{$K$-best algorithm}.  The $K$-best algorithm requires $O(NK^2 \log K)$ arithmetic operations. The maximum number of branches $K$ is free to be chosen.  For polynomial phase estimation we have found that setting $K=20N$ yields excellent results, and in this case the algorithm requires $O(N^3 \log N)$ operations.  Simulations in Section~\ref{sec:simulations} show that this algorithm produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for $N$ less than about 1000.

@@

Tables

\begin{table}[htbp]
\centering
\begin{tabular}{lrrrrr}
 & LSU & CPF & PHAF & HAF & ZW \\ \toprule
$N=99$ &  &  &  &  & \\ 
$N=199$ & 3 & 1 & 3 & 5 & 23 \\ 
$N=299$ &  &  &  & & \\ 
$N=399$ &  &  &  & & \\ 
$N=499$ &  &  &  & & \\ 
$N\to\infty$ & $-\infty$ & ? & ? & 3.77  & ? \\ \bottomrule
\end{tabular}
\caption{Threshold SNR in decibels for estimators with $m=3$}
\label{tablm3}
\end{table}

\begin{table}[htbp]
\centering
\begin{tabular}{lrrrrr}
 & LSU & CPF-HAF & PHAF & HAF & ZW \\ \toprule
$N=99$ &  &  &  &  & \\ 
$N=199$ & 12 & 10 & 12 & 18 & 103 \\ 
$N=299$ &  &  &  & & \\ 
$N=399$ &  &  &  & & \\ 
$N=499$ &  &  &  & & \\ 
$N\to\infty$ & $-\infty$ & $\approx 6.5$ & ?  & 16.6 & ? \\ \bottomrule
\end{tabular}
\caption{Threshold SNR in decibels for estimators with $m=5$}
\label{tablm3}
\end{table}

Figure~\ref{plot:bench} shows the running time of the LSU estimator computed using the $K$-best method versus the HAF estimator for polynomial phase signals of order 3.  The HAF estimator is computed using the fast Fourier transform and has complexity $O(N \log N)$.   Wtih $K=20N$ the LSU estimator computed using the $K$-best method requires $O(N^3 \log N)$ operations in the worst case.  The $K$-best method tends to be faster when the SNR is large.  This is due to the algorithm rapidly finding a nearest point, and subsequently terminating branches in its search tree early.  On the other hand, when the SNR is small, all branches in the tree tend to be searched.  Interestingly the LSU estimator computed using the $K$-best algorithm is comparable with the HAF when the SNR is large and when $N$ is approximately less than 100.  The LSU estimator can be computed in a reasonable amount of time (a few minutes) when $N=1000$.  The algorithm becomes prohibitively expensive for large $N$.  Benchmarks have been performed with polynomial phase signals of order larger than $3$.  The results are omitted due to space constraints.  The running time of the LSU estimator does not appear to change significantly as $m$ increases.  This is expected since a nearest point in a lattice of approximately the same dimension $N-m-1$ must be computed.  The complexity of the HAF estimator grows linearly with $m$.

@@

\nocite{OShea_refiningpps_2010,Amar_low_complex_polyphase_2010}
